---
title: "Literate Analysis with Ocean Water Samples"
subtitle: "coreR Session 05 — Section 5.7 Practice"
format:
  html:
    toc: true
    toc-depth: 3
execute:
  warning: false
  message: false
---

# Introduction

This Quarto document follows the **Section 5.7 practice exercise** from coreR *Session 05 (Literate Analysis with Quarto)*: building a simple **prose + code** analysis using real seawater chemistry data (`BGchem2008data.csv`).\
Source: <https://learning.nceas.ucsb.edu/2024-10-coreR/session_05.html>

## About the data

The seawater chemistry dataset used here is `BGchem2008data.csv` (downloaded previously in the course).

-   Date downloaded: August 8, 2025
-   Download source: <https://arcticdata.io/metacat/d1/mn/v2/object/urn%3Auuid%3A35ad7624-b159-4e29-a700-0c0770419941>

You will explore the nitrogen to phosphate ratio, which indicates the health of the ocean ecosystem. <https://beal-agulhas.earth.miami.edu/teaching/courses/lecture-five/index.html>

## Setup

Best practice: put all `library()` calls in one chunk near the top so collaborators can see what packages are required.

```{r}
library(here)
library(readr)
```

# Read in the data

Quarto runs R in the same way as a script: objects only exist after the chunks that create them have been run.

Using `here()` is a convenient way to build paths relative to your project root (e.g., your `.qmd` might live in `scripts/` while data lives in `data/`).

```{r}
bg_chem <- read_csv(here::here("data", "raw/BGchem2008data.csv"))
```

(Optional) Quick peek at the data:

```{r}
# Uncomment if you want more inspection output
 names(bg_chem)
 str(bg_chem)
 head(bg_chem)
 summary(bg_chem)
```

# Analysis

As the “analysis”, we’ll calculate a few simple summary statistics and examine how the nitrogen:phosphate values compare to the **Redfield ratio** concept.

There is a curious consistency that can occur in our planets ocean. The chemical make-up of algae floating in the water column tends to match the available nutrients in their environment. The Redford ratio is an elemental signature of Carbon, Nitrogen, and Phosphorus in the ratio 106:16:1. Scientists use the Redfield ratio to understand biological processes occurring the Ocean. When the available nutrients deviate from the Redfield ratio, it tells us something about how elements are circulating in the ocean and how living organisms may respond to global change.

## Calculate summary statistics

Compute mean nutrient values (using `na.rm = TRUE` so missing values don’t break the calculation).

```{r}
nitrate <- mean(bg_chem$NO3, na.rm = TRUE)
nitrite <- mean(bg_chem$NO2, na.rm = TRUE)
amm     <- mean(bg_chem$NH4, na.rm = TRUE)
phos    <- mean(bg_chem$P,   na.rm = TRUE)
bg_chem$N = bg_chem$NO3 + bg_chem$NO2 + bg_chem$NH4
```

## Calculate mean Redfield ratio

```{r}
ratio <- (nitrate + nitrite + amm) / phos
ratio
```

You can report results inline in Quarto like this:

The Redfield ratio for this dataset is approximately: `r round(ratio)`

## Plot Redfield ratio measurements

Finally, create a simple base R plot that shows the relationship between phosphate and the total nitrogen components.

```{r}
plot(
  bg_chem$P,
  bg_chem$NO2 + bg_chem$NO3 + bg_chem$NH4,
  xlab = "Phosphate (P)",
  ylab = "Total nitrogen (NO2 + NO3 + NH4)",
  main = "Ocean water samples: total N vs P"
)
```

There is a clear linear relationship between the concentration of phosphate to the concentration of total Nitrogen in ocean water. A linear regression would tell us if relationship is the Redfield ratio, but looking on this graph it does not seem that this is the 16:1 ratio. I wonder if the micrometer units require conversion to moles, and if the presence of oxygen and hydrogen in the total Nitrogen variable needs accounting for. Regardless of these details, I will assume that this is what the Redfield ratio looks like.

One interesting detail that is apparent is that there are outliers from the ratio, where phosphate occurs at lower than expected details. My class partner Esha noticed in that data that these outliers seem to be measurements taken at lower depths in the ocean. In the research by Martiny et al. published in 2013, it was noted that certain marine organisms are composed of different ratios of N:P. Specifically, cyanobacteria tend to be richer in phosphorus than Eukaryotes. Phosphorous is an important elemental in the construction of RNA and DNA. From this information, one might predict that cyanobacteria will be less numerous in environments with low concentration of phosphates.

# Conclusion

In this mini-analysis, we:

-   Loaded packages in a dedicated setup chunk
-   Read in `BGchem2008data.csv` using a project-root-relative path (`here()`)
-   Calculated simple nutrient means
-   Computed a mean nitrogen:phosphate ratio
-   Created a basic plot relating total nitrogen to phosphate

This is the core idea of **literate analysis**: combine narrative and code in one document so the workflow is easier to understand and reproduce.
